---
title: "2 Seurat Preprocessing"
output: html_notebook
---

```{r}
library(Seurat)
```

# Import and preparation
## Import and create Seurat objects
Import all 10X output files and create seurat objects
```{r}
# objects with "c" suffix are from replicate B, those without suffix from replicate A
# cutoff values set to min.cells = 3 and min.features = 200

regen.0hpa.data <- Read10X(data.dir = "data/matrix_files/0hpa_A/filtered_feature_bc_matrix")
trunk <- CreateSeuratObject(counts = regen.0hpa.data, project = "trunk", min.cells = 3, min.features = 200)
rm(regen.0hpa.data)

regen.12hpa.data <- Read10X(data.dir = "data/matrix_files/12hpa_A/filtered_feature_bc_matrix")
regen.12hpa <- CreateSeuratObject(counts = regen.12hpa.data, project = "Blastema_12hpa", min.cells = 3, min.features = 200)
rm(regen.12hpa.data)

regen.24hpa.data <- Read10X(data.dir = "data/matrix_files/24hpa_A/filtered_feature_bc_matrix")
regen.24hpa <- CreateSeuratObject(counts = regen.24hpa.data, project = "Blastema_24hpa", min.cells = 3, min.features = 200)
rm(regen.24hpa.data)

regen.48hpa.data <- Read10X(data.dir = "data/matrix_files/48hpa_A/filtered_feature_bc_matrix")
regen.48hpa <- CreateSeuratObject(counts = regen.48hpa.data, project = "Blastema_48hpa", min.cells = 3, min.features = 200)
rm(regen.48hpa.data)

regen.72hpa.data <- Read10X(data.dir = "data/matrix_files/72hpa_A/filtered_feature_bc_matrix")
regen.72hpa <- CreateSeuratObject(counts = regen.72hpa.data, project = "Blastema_72hpa", min.cells = 3, min.features = 200)
rm(regen.72hpa.data)

regen.0hpa.c.data <- Read10X(data.dir = "data/matrix_files/0hpa_B/filtered_feature_bc_matrix")
trunk.c <- CreateSeuratObject(counts = regen.0hpa.c.data, project = "trunk_c", min.cells = 3, min.features = 200)
rm(regen.0hpa.c.data)

regen.24hpa.c.data <- Read10X(data.dir = "data/matrix_files/24hpa_B/filtered_feature_bc_matrix")
regen.24hpa.c <- CreateSeuratObject(counts = regen.24hpa.c.data, project = "Blastema_24hpa_c", min.cells = 3, min.features = 200)
rm(regen.24hpa.c.data)

regen.48hpa.c.data <- Read10X(data.dir = "data/matrix_files/48hpa_B/filtered_feature_bc_matrix")
regen.48hpa.c <- CreateSeuratObject(counts = regen.48hpa.c.data, project = "Blastema_48hpa_c", min.cells = 3, min.features = 200)
rm(regen.48hpa.c.data)

regen.72hpa.c.data <- Read10X(data.dir = "data/matrix_files/72hpa_B/filtered_feature_bc_matrix")
regen.72hpa.c <- CreateSeuratObject(counts = regen.72hpa.c.data, project = "Blastema_72hpa_c", min.cells = 3, min.features = 200)
rm(regen.72hpa.c.data)
```

## give cells unique names
This avoids naming errors later on
```{r}
trunk <- RenameCells(trunk, add.cell.id = "trunk")
trunk.c <- RenameCells(trunk.c, add.cell.id = "trunkC")
regen.12hpa <- RenameCells(regen.12hpa, add.cell.id = "regen12hpa")
regen.24hpa <- RenameCells(regen.24hpa, add.cell.id = "regen24hpa")
regen.24hpa.c <- RenameCells(regen.24hpa.c, add.cell.id = "regen24hpaC")
regen.48hpa <- RenameCells(regen.48hpa, add.cell.id = "regen48hpa")
regen.48hpa.c <- RenameCells(regen.48hpa.c, add.cell.id = "regen48hpaC")
regen.72hpa <- RenameCells(regen.72hpa, add.cell.id = "regen72hpa")
regen.72hpa.c <- RenameCells(regen.72hpa.c, add.cell.id = "regen72hpaC")
```

## save pre-outlier removal object
Merge and store these samples before outlier removal for comparisons:
```{r}
allcells <- merge(trunk, y = c(
  trunk.c, regen.12hpa, regen.24hpa, regen.24hpa.c, regen.48hpa, regen.48hpa.c, regen.72hpa, regen.72hpa.c))

allcells$timepoint <- allcells$orig.ident
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_12hpa"] <- "12hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_24hpa"] <- "24hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_24hpa_c"] <- "24hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_48hpa"] <- "48hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_48hpa_c"] <- "48hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_72hpa"] <- "72hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="Blastema_72hpa_c"] <- "72hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="trunk"] <- "0hpa"
allcells$timepoint[allcells@meta.data$orig.ident=="trunk_c"] <- "0hpa"
saveRDS(allcells, file = "data/allcells_notcleaned.RDS")
```


# Outlier removal
Remove outlier cells based on counts and features (sample by sample for manual, visual inspection)
The cutoff values below are the ones used for the manuscript.

Set 
final <- FALSE 
to find the ideal cutoffs, then confirm and generate the new object by setting 
final <- TRUE

## Biological Replicate A
0hpa
```{r}
current.sample <- trunk
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 9000 & nFeature_RNA < 3000)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  trunk <- current.sample.sub
}
```

12hpa
```{r}
current.sample <- regen.12hpa
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 7500 & nFeature_RNA < 2700)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.12hpa <- current.sample.sub
}
```

24hpa
```{r}
current.sample <- regen.24hpa
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 17000 & nFeature_RNA < 5000)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.24hpa <- current.sample.sub
}
```

48hpa
```{r}
current.sample <- regen.48hpa
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 22000 & nFeature_RNA < 5000)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.48hpa <- current.sample.sub
}
```

72hpa
```{r}
current.sample <- regen.72hpa
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 35000 & nFeature_RNA < 6500)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.72hpa <- current.sample.sub
}
```

## Biological Replicate B

0hpa
```{r}
current.sample <- trunk.c
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 10000 & nFeature_RNA < 2500)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  trunk.c <- current.sample.sub
}
```

24hpa
```{r}
current.sample <- regen.24hpa.c
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 15000 & nFeature_RNA < 3100)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.24hpa.c <- current.sample.sub
}
```

48hpa
```{r}
current.sample <- regen.48hpa.c
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 15000 & nFeature_RNA < 3300)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.48hpa.c <- current.sample.sub
}
```

72hpa
```{r}
current.sample <- regen.72hpa.c
final <- TRUE
plot1 <- FeatureScatter(current.sample, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

current.sample.sub <- subset(current.sample, subset = nCount_RNA < 14000 & nFeature_RNA < 3500)
plot2 <- FeatureScatter(current.sample.sub, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

cells_before <- ncol(current.sample)
cells_after <- ncol(current.sample.sub)
cells_removed <- cells_before - cells_after
plot1
plot2

cells_before
cells_after
cells_removed

if(final == TRUE) {
  regen.72hpa.c <- current.sample.sub
}
```

# Main preprocessing steps
Following an approach found in 
https://github.com/technau/CellReports2022

```{r}
# merge objects
scd <- merge(trunk, y = c(trunk.c, regen.12hpa, regen.24hpa, regen.24hpa.c, regen.48hpa, regen.48hpa.c, regen.72hpa, regen.72hpa.c), merge.data = F)

# normalize (across dataset)
scd <- NormalizeData(scd, normalization.method = "LogNormalize", scale.factor = 10000)

# all samples are also individually normalized here; then variable features are selected and written into a list. 
list = NULL
regen.vargenes.list <- SplitObject(scd, split.by = "orig.ident")

for (i in 1:length(regen.vargenes.list)) {
    regen.vargenes.list[[i]] <- FindVariableFeatures(regen.vargenes.list[[i]], selection.method = "vst", nfeatures = 2000, verbose = FALSE)  
}

# collate into a single list of unique features and import into the object:
for (i in 1:length(regen.vargenes.list)) {
  x <- regen.vargenes.list[[i]]@assays$RNA@var.features
  list=c(list,x)}
var.features.list=unique(list)

# How many variable features does the dataset contain across all samples?
length(var.features.list)

# store them in the variable features slot of the merged object:
scd@assays$RNA@var.features = var.features.list

# now scale data, based on sample of origin (centers at 0, variance scaled to 1 so all variable genes are weighted equally downstream)
scd <- ScaleData(scd, features = scd@assays$RNA@var.features, split.by = "orig.ident", do.scale = T, do.center = T)

# calculate principal components; calculate umap; find neighbors
scd <- RunPCA(scd, npcs = 50)
scd <- RunUMAP(scd, reduction = "pca", dims = 1:50)
scd <- FindNeighbors(scd, reduction = "pca", dims = 1:50)

# cluster the dataset
scd <- FindClusters(scd, resolution = 0.5)
scd <- SetIdent(scd, value = "RNA_snn_res.0.5")

saveRDS(scd, file = "data/scd.RDS")
```
