---
title: "03 additional sc processing and subclustering"
output: html_notebook
---

```{r}
library(Seurat)
library(tidyverse)
scd <- readRDS(file = "data/scd.RDS")
```

# create log-transformed data slot 
standard data transformation and normalisation workflows scale the counts by sample (i.e. cell) size. In case of genuinely larger cells with higher RNA content, this might quench biologically important signal (see ref. )
We therefore generated logtransformed (log1p) data without scaling to alleviate this issue. The log transformation is required to reduce the overall scale range/ visually better.

```{r}
log_assay <- CreateAssayObject(data = scd@assays$RNA@data)
scd[["logdata"]] <- log_assay
scd@assays$logdata@data <- log1p(scd@assays$RNA@counts)
```

# add metadata information
```{r}
# sampling timepoint
scd$timepoint <- scd$orig.ident
scd$timepoint[scd@meta.data$orig.ident=="Blastema_12hpa"] <- "12hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_24hpa"] <- "24hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_24hpa_c"] <- "24hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_48hpa"] <- "48hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_48hpa_c"] <- "48hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_72hpa"] <- "72hpa"
scd$timepoint[scd@meta.data$orig.ident=="Blastema_72hpa_c"] <- "72hpa"
scd$timepoint[scd@meta.data$orig.ident=="trunk"] <- "0hpa"
scd$timepoint[scd@meta.data$orig.ident=="trunk_c"] <- "0hpa"

table(scd$timepoint)

# sampling run
scd$sample <- scd$orig.ident
scd$sample[scd@meta.data$orig.ident=="Blastema_12hpa"] <- "A"
scd$sample[scd@meta.data$orig.ident=="Blastema_24hpa"] <- "A"
scd$sample[scd@meta.data$orig.ident=="Blastema_24hpa_c"] <- "B"
scd$sample[scd@meta.data$orig.ident=="Blastema_48hpa"] <- "A"
scd$sample[scd@meta.data$orig.ident=="Blastema_48hpa_c"] <- "B"
scd$sample[scd@meta.data$orig.ident=="Blastema_72hpa"] <- "A"
scd$sample[scd@meta.data$orig.ident=="Blastema_72hpa_c"] <- "B"
scd$sample[scd@meta.data$orig.ident=="trunk"] <- "A"
scd$sample[scd@meta.data$orig.ident=="trunk_c"] <- "B"

table(scd$sample)

# injury state
scd <- SetIdent(scd, value = "timepoint")
scd$isinjured <- scd@meta.data$timepoint
injury.ident <- c("uninjured", "injured", "injured", "injured", "injured")
names(injury.ident) <- levels(scd)
scd <- RenameIdents(scd, injury.ident)
scd$isinjured <- scd@active.ident
scd <- SetIdent(scd, value = "RNA_snn_res.0.5")

table(scd$isinjured)
```

# Subclustering
## Subsetting + re-run of preprocessing
```{r}
# store clusters to new slot before overriding them when re-processing again
scd$res0.5.clusters <- scd$RNA_snn_res.0.5

# subset clusters 0 and 1
sub.c0.c1 <- subset(scd, idents = c(0, 1))

# clean their data and re-run preprocessing
sub.c0.c1 <- DietSeurat(sub.c0.c1, scale.data = F) %>%
  NormalizeData(normalization.method = "LogNormalize", scale.factor = 10000) %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 3000) %>%
  ScaleData() %>%
  RunPCA(npcs = 50) %>%
  RunUMAP(reduction = "pca", dims = 1:50) %>%
  FindNeighbors(reduction = "pca", dims = 1:50) %>%
  FindClusters(resolution = 0.5)

# merge clusters based on shared expression of SAZ-related genes
newidents <- c(0,1,2,12,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,12,20)
names(newidents) <- levels(sub.c0.c1)
sub.c0.c1 <- RenameIdents(sub.c0.c1, newidents)

# generate log1p transformed counts
log_assay <- CreateAssayObject(data = sub.c0.c1@assays$RNA@data)
sub.c0.c1[["logdata"]] <- log_assay
sub.c0.c1@assays$logdata@data <- log1p(sub.c0.c1@assays$RNA@counts)
DefaultAssay(sub.c0.c1) <- "logdata"

save(file = "data/c0c1_subset.RData", sub.c0.c1)
```

## label transfer to main object
```{r}
# store subcluster information in new metadata slot called subclustered
# only transfer relevant subclusters
subsub.c7.c12 <- subset(sub.c0.c1, idents = c("7", "12"))

scd$subclustered <- as.character(Idents(scd))
scd$subclustered[Cells(subsub.c7.c12)] <- paste("sub", Idents(subsub.c7.c12), sep = "")
saveRDS(scd, file = "data/scd.RDS")
```
