---
title: "06_cytotrace"
output: html_notebook
---

```{r}
library(Seurat)
library(CytoTRACE)
library(openxlsx)
library(tidyverse)
library(SCpubr)
library(ggplot2)

scd <- readRDS("data/scd.RDS")
load("data/xloc_blast_hits.RData")

# if you have run this before, you can read in the cytotrace data using the line below.
# cytotrace.all <- readRDS("data/cytotrace_all.RDS")

```

# Prepare data
We merge some already connected clusters of similar cell types first, as cytotrace is calculated per cluster.

```{r}
clusnames.new <- as.character(
  c(0:2,
    "3_6_12",
    "4_16",
    5,
    "3_6_12",
    7,
    "8_14",
    9:11,
    "3_6_12",
    13,
    "8_14",
    15,
    "4_16",
    17:37)
  )

names(clusnames.new) <- levels(scd)
scd <- RenameIdents(scd, clusnames.new)

scd$clus.merged <- scd@active.ident
scd <- SetIdent(scd, value = "clus.merged")
```

# Run CytoTrace
```{r, eval=FALSE}
cytotrace.all <- list()
clusters <- unique(scd$clus.merged)
for (clus in clusters) {
  cluster.sub <- subset(scd, idents = clus) # subset one cluster
  cluster.matrix <- as.matrix(GetAssayData(cluster.sub, 
                                           assay = "RNA", 
                                           slot = "counts")) # extract counts
  cytotrace.data <- CytoTRACE(cluster.matrix, enableFast = TRUE, ncores = 8) # run CT on 8 cores
  cytotrace.all[[clus]] <- cytotrace.data # store result to list
}
 
saveRDS(cytotrace.all, file = "data/cytotrace_all.RDS") # saves cytotrace results
```

# Transfer cytotrace to seurat metadata
```{r}
clusters <- unique(scd$clus.merged)

# loop through list of cytotrace results, store per cluster
for (cluster in clusters) {
  cells_in_cluster <- WhichCells(scd, idents = cluster)
  cytotrace_values <- cytotrace.all[[as.character(cluster)]]$CytoTRACE
  scd@meta.data[cells_in_cluster, "cytotrace"] <- cytotrace_values
}

# quick glance at the data
FeaturePlot(scd, features = "cytotrace")
DimPlot(scd, group.by = "timepoint", shuffle = T)
```

store new scd with cytotrace
```{r}
scd <- SetIdent(scd, value = "RNA_snn_res.0.5")
save(scd, file = "data/scd.RData")
```

# Genes correlated with CytoTRACE
i.e. up/downregulated in cells of higher developmental potential.
```{r}
df.list <- list()
for(cluster in names(cytotrace.all)) {
  ct.gene <- cytotrace.all[[cluster]]$cytoGenes
  gene_names <- names(ct.gene)
  df <- tibble(gene_name = gene_names, !!cluster := ct.gene)
  df.list[[cluster]] <- df
}

ct.genes.all <- reduce(df.list, full_join, by = "gene_name")
```

annotate with BLAST hits
```{r}
ct.genes.ann <- ct.genes.all %>%
  left_join(xloc.annotations, by = c("gene_name" = "XLOC"))
```

annotate with gene IDs for GO terms
(GO term conversion information in methods section of manuscript)
```{r}
GO.droso <- read.delim(file = "/Users/alexanderstockinger/Informatics/2024_atlas_playground/data/GOterms_pdu_2024/XLOC2up_Droso.tab", header = F, col.names = c("xloc", "go_droso"))
GO.droso$xloc <- sub("_", "-", GO.droso$xloc)

GO.hsap <- read.delim(file = "/Users/alexanderstockinger/Informatics/2024_atlas_playground/data/GOterms_pdu_2024/XLOC2up_Homo.tab", header = F, col.names = c("xloc", "go_human"))
GO.hsap$xloc <- sub("_", "-", GO.hsap$xloc)

ct.genes.ann <- ct.genes.ann %>%
  left_join(GO.droso, by = c("gene_name" = "xloc")) %>%
  left_join(GO.hsap, by = c("gene_name" = "xloc"))
```

Reorder columns
```{r}
clusnames.un <- as.character(
  c("gene_name",
    "hits",
    "go_droso",
    "go_human",
    0:2,
    "3_6_12",
    "4_16",
    5,
    7,
    "8_14",
    9:11,
    13,
    15,
    17:37)
  )

ct.genes <- ct.genes.ann[, clusnames.un]
```

save and export
```{r}
write.xlsx(ct.genes, file="data/cytotrace_all_genes.xlsx", colNames = TRUE)
save(ct.genes, file = "data/genes.with.goterms.RData")
```

# Plotting
## Main umap with cytotrace
```{r}
p1 <- SCpubr::do_FeaturePlot(scd, 
                             features = "cytotrace", order = F,
                             pt.size = 0.1, legend.position = "bottom", legend.type = "colorbar") + 
  theme(plot.margin =unit(c(0,0,0,0), "pt"), aspect.ratio = 1)
p1

ggsave(p1, filename = "plots/cytotrace/cytotrace_umap_c.png",
       width = 15, height = 12, units = "cm")
```

## subclusters with cytotrace
transfer cytotrace info:
```{r}
cells_in_cluster <- WhichCells(scd, idents = cluster)
  cytotrace_values <- cytotrace.all[[as.character(cluster)]]$CytoTRACE
  scd@meta.data[cells_in_cluster, "cytotrace"] <- cytotrace_values
  
  
cells.use <- WhichCells(scd, idents = c("0", "1"))
sub.c0.c1$cytotrace <- scd$cytotrace[cells.use]
```

```{r}
# load(file = "data/c0c1_subset.RData")
SCpubr::do_DimPlot(sub.c0.c1, pt.size = 0.2, legend.position = "right") + 
  theme(plot.margin =unit(c(0,0,0,0), "pt"), aspect.ratio = 1)

SCpubr::do_FeaturePlot(scd, features = "cytotrace",
                       pt.size = 0.2, legend.position = "right") + 
  theme(plot.margin =unit(c(0,0,0,0), "pt"), aspect.ratio = 1)
```



